### 2019-08-01 paper/abstract.tex
outlining three arguments

Due to a lack of precise ages and rotation periods for old main-sequence
stars, the rotational evolution of FGKM dwarfs is mostly unconstrained
after around 2-3 billion years.
% This limit is set by the oldest open clusters whose members have precise light
% curves, and therefore photometric rotation periods.
Recent evidence suggests that the relationship between rotation period and
color/temperature/mass does not remain constant over large timescales,
contradicting the long-held assumption that underlies many gyrochronology
relations.
% For example, old clusters appear to have a flatter relationship between
% rotation period and color than young clusters.
In this work we use the kinematic ages of populations of K dwarfs to reveal
the evolving relationship between rotation period and effective temperature.
We find that groups of stars, selected to be the same age, do {\it not} have
the same velocity dispersion at all temperatures; cooler stars have greater
velocity dispersions than hotter stars.
Two different scenarios could be responsible for producing this observation:
either the rotation period-color relation changes over time, or stars of
different masses experience different rates of dynamical heating.
We show however, that if mass-dependent heating is responsible, the
differential heating rate is not constant over logarithmic time intervals
which implies that different heating processes must be acting before and after
ages of around 1 Gyr.
Perhaps a more likely scenario is that the relation between rotation period
and color changes over time, undergoing a particularly sharp transition at
around 1 Gyr.
If we assume that mass-dependent heating does not affect the data, our results
indicate that the slope of the period-color relation actually {\it inverts}:
at young ages early K stars rotate more rapidly than late K stars, but beyond
$\sim$ 2 Gyr they rotate more slowly.
% (similar to the age of the rotation gap and the 1.1 Gyr NGC 6811
% cluster, shown to have a different period-color relation to Praesepe).
% We find that the ages of field stars with measured rotation periods do not
% exceed 4-6 Gyrs, resolving a long-standing degeneracy between age and magnetic
% braking efficiency.
% We also show that stars rotating just below the rotation period gap are
% dynamically young, ruling out the possibility that the rotation period gap is
% caused by incorrect period measurements or binary companions.
% The main result of this paper could be produced by either an evolving
% period-color relationship {\it or} mass-dependent dynamical heating.
% Although we provide some evidence to suggest that the former effect is
% dominant, we cannot be sure and therefore caution that the results presented
% here are subject to this caveat.

>>>>

Due to a lack of precise ages and rotation periods for old main-sequence
stars, the rotational evolution of FGKM dwarfs is mostly unconstrained
after around 2-3 billion years.
However, recent evidence suggests that the relationship between rotation
period and color/temperature/mass does not remain constant over large
timescales, contradicting a long-held assumption that underlies many
gyrochronology relations.
In this work we use the kinematic ages of populations of K dwarfs to reveal
the evolving relationship between rotation period and effective temperature.
We find that groups of stars, selected to be the same age, do {\it not} have
the same velocity dispersion at all temperatures; cooler stars have greater
velocity dispersions than hotter stars.
Two different scenarios could be responsible for producing this observation:
either the rotation period-color relation changes over time, or stars of
different masses experience different rates of dynamical heating.
% We show however, that if mass-dependent heating is responsible, the
% differential heating rate is not constant over logarithmic time intervals
% which implies that different heating processes must be acting before and after
% ages of around 1 Gyr.
% Perhaps a more likely scenario is that the relation between rotation period
% and color changes over time, undergoing a particularly sharp transition at
% around 1 Gyr.
If we assume that mass-dependent heating does not affect the data, our results
indicate that the slope of the period-color relation actually {\it inverts}:
at young ages early K stars rotate more rapidly than late K stars, but beyond
$\sim$ 2 Gyr they rotate more slowly.

% (similar to the age of the rotation gap and the 1.1 Gyr NGC 6811
% cluster, shown to have a different period-color relation to Praesepe).
% We find that the ages of field stars with measured rotation periods do not
% exceed 4-6 Gyrs, resolving a long-standing degeneracy between age and magnetic
% braking efficiency.
% We also show that stars rotating just below the rotation period gap are
% dynamically young, ruling out the possibility that the rotation period gap is
% caused by incorrect period measurements or binary companions.

### 2019-08-01 paper/conclusion.tex
outlining three arguments

We examined the rotation period-color component of the \citep{angus2019}
gyrochronology relation and found that groups of stars, selected to be the
same age using this relation, do {\it not} have the same velocity dispersion
across all temperatures.
This indicates that {\it either} the \citep{angus2019} period-color relation
is wrong, {\it or} that mass-dependent dynamical heating is affecting this
data set.
We found that the exponent of the AVR relation, calibrated using F and G stars
in the Geneva-Copenhagen Survey, with ages estimated via isochrone-fitting, is
greater than the exponent we find for the AVR relation fit to the most massive
K dwarfs in our sample.
Since mass-dependent dynamical heating should lead to a larger AVR exponent
for lower-mass stars, this suggests that mass-dependent heating is not
strongly affecting our sample.
However, we used velocity in galactic latitude, \vb, not \vz, to calculate our
AVRs, and while \vb\ is a close approximation to \vz\ at low galactic
latitudes, it is not identical.
Indeed, the \vz\ AVR should be larger than the \vb\ AVR, since orbital
scattering is expected to have a strongest effect in the vertical direction.
For the 290 stars in our sample with \gaia\ radial velocity measurements we
measured the \vz\ AVR exponent to be larger than the \vb\ AVR exponent.

For now, we cannot draw a strong conclusion regarding the origin of the
observed increased velocity dispersion for stars of lower masses since either
an incorrectly calibrated gyrochronology relation, or mass-dependent dynamical
heating could be the cause.

>>>>

We found that groups of K dwarfs, selected to be coeval using the
\citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
dispersion across all temperatures.
This indicates that {\it either} the \citep{angus2019} period-color relation
is wrong, {\it or} that mass-dependent dynamical heating is affecting this
data set.
We presented three pieces of evidence that suggest mass-dependent heating
cannot be responsible for all of the increasing velocity dispersion.

\begin{itemize}

\item{No strong evidence for mass-dependent heating has been found among M, L
and T-type stars.
These low-mass stars live longer than the current age of the Universe and
therefore do not suffer from a mass-age degeneracy \citep{faherty2009}.}

\item{Literature measurements of the (\vz) AVR exponent are slightly higher
than the value measured from our data set, but calculated using more massive
stars.
Mass-dependent dynamical heating should however produce a {\it smaller} AVR
exponent for higher-mass stars.
We acknowledge that the stars in the \mct\ sample, analyzed in this work, are
subject to different selection biases than those used to calculate AVRs in the
literature \citep{holmberg2009, aumer2009, yu2018}, so drawing direct
comparisons between them is not a robust test.}

\item{The differences in the (\vb) AVR exponent between the hottest and
coolest K dwarfs in our sample are extremely large, spanning values between
0.3 and 0.65, yet the difference in mass is less than a factor of two.
Such a large spread in AVR exponents cannot be explained by mass-dependent
heating alone.}

\end{itemize}

\racomment{Words about the significance of an inverted period-color relation.}

### 2019-08-01 paper/intro.tex
outlining three arguments

Stars are thought to be born in the thin disk of the Milky Way, orbiting the
center of the galaxy with a low out-of-plane, or vertical, velocity ($W$, or
$v_z$).
Young stars have relatively small vertical velocities, but gain momentum in
the vertical direction over time.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds are thought to play an important role
\racomment{(citations)}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
% Stellar velocities have a history of being used as an age proxy, with several
% notable examples within stellar astronomy \citep[\eg][]{faherty2009,
% west2011}.
Since the age-velocity dispersion relations (AVRs) are themselves still
actively being calibrated, it is difficult to directly compare gyrochronal
ages with kinematic ones.
However, regardless of the exact relation between velocity dispersion and
stellar age, it is expected to be a monotonic relationship, therefore velocity
dispersion can be used effectively to {\it rank} groups of stars by age.
\racomment{Say some words here to prime the reader for the upcoming caveats.}

% Vertical {\it actions} are better age indicators than velocities, because
% actions are calculated by integrating angular momentum over the Milky Way's
% potential, and are therefore position invariant -- \eg\ a star will have the
% same action at periapsis and apoapsis.
% In contrast, orbital velocities are different at periapsis and apoapsis -- so
% in this sense, {\it actions} are the natural quantities to use as age proxies.
Vertical action is a better age indicator than vertical velocity, however both
vertical action ($J_z$) and vertical velocity (\vz/W) can only be calculated
with full 6-dimensional position and velocity information, yet most stars with
measured rotation periods do not have radial velocity measurements because
they are relatively faint \kepler\ targets ($\sim$11th-18th magnitudes).
For this reason, we used an alternative age proxy: velocity in the direction
of galactic latitude, \vb.
The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
\vb\ is a close approximation to \vz.

@@

In this paper we test the \citet{angus2019} gyrochronology relation, which
was calibrated using the period-color relation of Praesepe and the period-age
relation of Praesepe and the Sun.
It is calibrated to \gaia\ \gcolor\ color.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
Although only calibrated using Praesepe and the Sun, this gyrochronology
relation was tested on NGC 6819, a 2.5 Gyr cluster, for which it predicted
accurate ages.
In addition, the period-color relation of Praesepe is nearly identical to the
period-color relation of the Hyades, a cluster of around the same age: $\sim$
650 years \citep{douglas2019}.
However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1
Gyr cluster \citep{curtis2019}.
The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
appears as though the K stars have `stalled' -- their spin-down has been
halted.
The \citet{angus2019} gyrochronology relation assumes that the rotation
period-color relation of Praesepe is applicable to stars of all ages: the same
polynomial relation fit to Praesepe is used to describe the period-color
relation for all stars.
However, the NGC 6811 cluster suggests that this assumption does not hold past
around 1 Gyr.
If the period-color relation of the \citet{angus2019} gyrochronology model
{\it were} a perfect model for the rotational evolution of stars, then groups
of stars selected to be similar ages using this relation should fall on an
isochrone on a CMD, an isochrone in period-color space, and an
isochrone in velocity dispersion-color space.
Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
variations in metallicity and extinction, blur out the main sequence enough
that differences between stars in different age and rotation bins are not
easily discernable on the CMD, although there is still a general age and
rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
\ref{fig:age_gradient}.
For this reason, we chose to use {\it kinematic} `isochrones', instead of
magnitude and color-based isochrones: although kinematics as an age indicator
is not necessarily as well calibrated to an absolute age scale as CMD
position, it can be very sensitive to {\it differences} in the ages of stellar
populations.

>>>>

Stars are thought to be born in the thin disk of the Milky Way, orbiting the
center of the galaxy with a low out-of-plane, or vertical, velocity ($W$, or
$v_z$).
Young stars have relatively small vertical velocities, but gain momentum in
the vertical direction over time.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds and spiral arms are thought to play an important
role \racomment{(citations)}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
% Stellar velocities have a history of being used as an age proxy, with several
% notable examples within stellar astronomy \citep[\eg][]{faherty2009,
% west2011}.
Since the age-velocity dispersion relations (AVRs) are themselves still
actively being calibrated, it is difficult to directly compare gyrochronal
ages with kinematic ones.
However, regardless of the exact relation between velocity dispersion and
stellar age, it is expected to be a monotonic relationship, therefore velocity
dispersion can be used effectively to {\it rank} groups of stars by age.

% Vertical {\it actions} are better age indicators than velocities, because
% actions are calculated by integrating angular momentum over the Milky Way's
% potential, and are therefore position invariant -- \eg\ a star will have the
% same action at periapsis and apoapsis.
% In contrast, orbital velocities are different at periapsis and apoapsis -- so
% in this sense, {\it actions} are the natural quantities to use as age proxies.
Vertical action is a better age indicator than vertical velocity, however both
vertical action ($J_z$) and vertical velocity (\vz/W) can only be calculated
with full 6-dimensional position and velocity information.
Unfortunately, most stars with measured rotation periods do not have radial
velocity (RV) measurements because they are relatively faint \kepler\ targets
($\sim$11th-18th magnitudes).
For this reason, we used an alternative age proxy: velocity in the direction
of galactic latitude, \vb.
The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
\vb\ is a close approximation to \vz.

@@

In this paper, we tested the \citet{angus2019} gyrochronology relation, which
was calibrated using the period-color relation of Praesepe (in \gaia\ \gcolor\
color) and the period-age relation of Praesepe and the Sun.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
Although only calibrated using Praesepe and the Sun, this gyrochronology
relation was tested on NGC 6819, a 2.5 Gyr cluster, for which it predicted
accurate ages.
In addition, the period-color relation of Praesepe is nearly identical to the
period-color relation of the Hyades, a cluster of around the same age: $\sim$
650 years \citep{douglas2016, douglas2017, rebull2017}.
However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1
Gyr cluster \citep{curtis2019}.
The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
appears as though the K stars have `stalled' -- their spin-down has been
halted.
The \citet{angus2019} gyrochronology relation assumes that the rotation
period-color relation of Praesepe is applicable to stars of all ages: the same
polynomial relation fit to Praesepe is used to describe the period-color
relation for all stars.
However, the NGC 6811 cluster suggests that this assumption does not hold past
around 1 Gyr.
If the period-color relation of the \citet{angus2019} gyrochronology model
{\it were} a perfect model for the rotational evolution of stars, then groups
of stars selected to be similar ages using this relation should fall on an
isochrone on a CMD, an isochrone in period-color space, and an
isochrone in velocity dispersion-color space.
Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
variations in metallicity and extinction, blur out the main sequence enough
that differences between stars in different age and rotation bins are not
easily discernable on the CMD, although there is still a general age and
rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
\ref{fig:age_gradient}.
For this reason, we chose to use {\it kinematic} `isochrones', instead of
magnitude and color-based isochrones: although kinematics as an age indicator
is not necessarily as well calibrated to an absolute age scale as CMD
position, it can be very sensitive to {\it differences} in the ages of stellar
populations.

### 2019-08-01 paper/results.tex
outlining three arguments

Mass-dependent dynamical heating could occur because lower-mass stars
experience greater velocity changes when gravitationally perturbed.
However, mass-dependent orbital heating has not yet been unambiguously
observed in the galactic disk because of the strong anti-correlation between
stellar mass and stellar age.
Less massive stars do indeed have larger velocity dispersions, however they
are also older on average.
This mass-age degeneracy is highly reduced in M dwarfs because their
main-sequence lifetimes are longer than the age of the Universe, however no
evidence for mass-dependent heating has been detected in these low mass stars
\citep{faherty2009}.

To investigate further, we calculated the exponent of the AVR for each
temperature bin in figure \ref{fig:age_cut}.
If mass-dependent heating is strongly affecting our data, the exponent of the
AVR should increase with decreasing effective temperature, \ie\ the heating
rate should be greater for lower-mass stars.
Figure \ref{fig:AVR_exponent} shows the AVR exponent does indeed increase with
decreasing effective temperature, however, the AVR exponent calculated using
more massive stars from the Geneva Copenhagen Survey (GCS)
\citep{holmberg2009} is also shown on this plot as a dashed horizontal line.
Most GCS stars are F and G type: 1-3 times more massive than the K dwarfs used
in our study.
If mass-dependent orbital heating is the main cause the rise in velocity
dispersion with decreasing \teff\ seen in figure \ref{fig:age_cut}, then the
AVR exponent of the more massive GCS stars should fall {\it below} the AVR
exponent for the most massive stars in our sample.
In other words, the dynamical heating rate should be lowest for the highest
mass stars and highest for the lowest mass stars.
\begin{figure}
  \caption{
The exponent of the AVR as a function of effective temperature.
    The AVR exponent of \citet{holmberg2009}, calculated using a more massive
    sample of stars, is shown as a horizontal dashed line.
}
  \centering
    \includegraphics[width=1\textwidth]{AVR_exponent}
\label{fig:AVR_exponent}
\end{figure}
% nordstrom2004, jorgensen2005,
These AVR exponents were calculated with different populations of stars,
subject to different selection effects, so directly comparing them comes with
some risk.
The most important difference is that the GCS AVR is calculated using \vz,
however our AVRs are calculated using \vb\ since most stars in our sample do
not have radial velocities.
The median galactic latitude of these \kepler\ stars is 12.3\degrees, with
latitudes ranging from 5.5\degrees\ to 21.4\degrees, so \vb\ is a relatively
close approximation to \vz.
Only 290 out of the 6820 stars in our sample, plotted in color in figure
\ref{fig:age_cut}, have \gaia\ radial velocities, however we used the ones
that do to compare the \vz\ AVR to the \vb\ AVR, across all temperatures
between 3500 and 5000 K.
For stars with gyrochronal ages between 0.5 and 4.5 Gyr, we measured a \vz\
AVR exponent of 0.519 $\pm$ ... and a \vb\ AVR exponent of 0.514 $\pm$... .
The similarity of these two AVR exponents suggests that directly comparing
these two quantities {\it may} be a reasonable approach.
For now, given the large difference between the (\vz) AVR exponent of the GCS
sample and the (\vb) AVR exponent of the hottest stars in our sample, we
assume that the increased velocity dispersion at cooler temperatures is mostly
caused by incorrect age-grouping due to an incorrect period-color relation at
old ages, and that any mass-dependent heating, while it may contribute at a
low level to this result, is not the dominant driver.
However, differentiating the effects of mass-dependent heating and the shape
of the gyrochronology relations is certainly warranted in a follow-up study.

The Praesepe-calibrated gyrochronology relation seems to accurately predict
the relative ages of {\it young} field stars.
This is shown by the consistent velocity dispersion of young stars as a
function of temperature in figure \ref{fig:age_cut}.
To our knowledge, no gyrochronology relation has never been demonstrated to
correctly predict ages (either relative or absolute) for such cool or such
young field stars.
This is not particularly remarkable however, since these young stars are a
similar age to the Praesepe cluster, which was used to calibrate the
period-color relation.
Figure \ref{fig:age_comparison} shows the ages of star groups, predicted with
the \citet{angus2019} gyrochronology relation, compared with the ages of star
groups, predicted with the \citet{holmberg2009} AVR.
\citet{holmberg2009} only provide an exponent (0.53), not an intercept, for the
relation between logarithmic age (in Gyr) and logarithmic, \vz\ velocity
dispersion (in \kms), so we fit the intercept to the mean \vb\ velocity
dispersion across all temperatures, of our sample.
\begin{figure}
  \caption{
      Ages predicted by the \citet{holmberg2009} \vz\ AVR, against ages
    predicted by the \citet{angus2019} gyrochronology relation, for stars in
    different temperature ranges.
Points are colored by the effective temperature in the center of the \teff\
bin.
The dashed line shows the y=x relation.
Predicted kinematic ages fan out over gyrochronal time, either because the AVR
    is temperature-dependent or because the gyrochronology relations have an
    age-dependent period-color relation.
}
  \centering
    \includegraphics[width=1\textwidth]{age_comparison}
\label{fig:age_comparison}
\end{figure}
Figure \ref{fig:age_comparison} shows a comparison between ages predicted by
the \citet{holmberg2009} \vz\ AVR (using \vb\ as a proxy for \vz) and ages
predicted by the \citet{angus2019} gyrochronology relation, for stars in
different temperature ranges.
The predicted kinematic ages fan out over gyrochronal time, either because the
AVR is temperature-dependent or because the gyrochronology relations have an
age-dependent period-color relation.
Figure \ref{fig:age_comparion} also shows that the AVR seems to break down for
the youngest age bin, 0.5 Gyr, which has a lower velocity dispersion than
expected, given its age.
% Alternatively, the ages of these young stars could be under-predicted by
% gyrochronology.
% These young stars have similar rotation periods to the Praesepe cluster, which
% was used to calibrate the gyrochronology relation used in this analysis.
% The ages of these young stars should, therefore, be the most accurate of the
% entire sample.
% However, they are tied to the age of the Praesepe cluster, whose age is not
% accurately known.
% \citet{angus2019} adopted an age for Praesepe of 650 Myrs, however the
% kinematic age-prediction for these stars is around 400 Myrs.
% Other studies have assigned a much older age of 800 Myrs to Praesepe
% \citep{brandt2015}.
% If the age assumed for Praesepe in the \citet{angus2019} gyrochronology
% calibration was inaccurate, the entire age {\it scale} would be incorrect.
% However, this is not what figure \ref{fig:age_comparion} shows -- it shows
% that the {\it relative} age of these youngest stars are under-predicted by
% kinematics or over-predicted by gyrochronology.

We applied the same analysis described at the beginning of this section to
groups of stars selected using rotation period groups, instead of age groups.
The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with
stars in different period ranges plotted in different colors.
The bottom panel shows the velocity dispersion of each group as a function of
effective temperature.
\begin{figure}
  \caption{
This figure is similar to figure \ref{fig:age_cut}, with stars divided into
    period groups rather than age groups.
The velocity dispersion is more constant across effective temperatures for the
    most slowly rotating stars, compared to the stars selected with the
    \citet{angus2019} gyrochronology model, indicating that the gyrochronology
    models flatten out, and possibly even invert, at old ages.
}
  \centering
    \includegraphics[width=1\textwidth]{period_cut}
\label{fig:period_cut}
\end{figure}
Once again, the velocity dispersion increases with rotation period overall.
For the most rapidly rotating groups of stars, velocity dispersion decreases
with \teff.
This is expected given the positively sloped period-color relation of Praesepe
and other young clusters: late K dwarfs rotate more slowly than early K dwarfs
of the same age.
Between 15 and 25 day rotation periods, the temperature dependence of the
velocity dispersion starts to disappear, indicating that the period-color
relation becomes flat: late K dwarfs rotate at the same rate as early K dwarfs
of the same age.
At long rotation periods, the velocity dispersion still increases with
effective temperature, although the increase is much more modest than in
figure \ref{fig:age_cut}.
This suggests that the period-color relation may have a negative slope at old
ages: late K dwarfs may rotate {\it more} rapidly than early K dwarfs.
This would be a paradigm shift for gyrochronology, stellar spin-down rate is
thought to be directly tied to magnetic field strength.
The deeper convection zones of later type stars generate stronger magnetic
fields which {\it should} lead to more efficient angular momentum loss.

>>>>

Mass-dependent dynamical heating could occur because lower-mass stars
experience greater velocity changes when gravitationally perturbed.
However, mass-dependent orbital heating has not yet been unambiguously
observed in the galactic disk because of the strong anti-correlation between
stellar mass and stellar age.
Less massive stars do indeed have larger velocity dispersions, however they
are also older on average \racomment{(citations)}.
This mass-age degeneracy is highly reduced in M dwarfs because their
main-sequence lifetimes are longer than the age of the Universe, however no
evidence for mass-dependent heating has been detected in these low mass stars
\citep{faherty2009}.

The \vb\ AVR is not directly comparable to the \vz\ AVR, so, to draw a
comparison with results from the literature, we calculated a \vz\ AVR for
the subset of 290 stars in our sample with \gaia\ RVs.
We measured an AVR exponent of 0.47 $\pm$ 0.1, which falls within the range of
values (0.45-0.53) reported from measurements of F and G stars in the Solar
neighborhood \citep{holmberg2009, aumer2009, aumer2016}.
It is a little lower than the values of 0.56 $\pm$ 0.14 (for low-z) and 0.51
$\pm$ 0.15 (for high-z stars), calculated using \LAMOST\ \racomment{(LAMOST
citation)} K giants \citep{yu2018}.
K giants are more massive than K dwarfs \racomment{(how much more massive?)},
so this slightly higher value contradicts the mass-dependent heating
hypothesis.
\racomment{Add some words about selection functions...}

To investigate further, we calculated the exponent of the (\vb) AVR for each
temperature bin in figure \ref{fig:age_cut}.
\begin{figure}
  \caption{
The exponent of the AVR as a function of effective temperature.
    The AVR exponent of \citet{holmberg2009}, calculated using a more massive
    sample of stars, is shown as a horizontal dashed line.
}
  \centering
    \includegraphics[width=1\textwidth]{AVR_exponent}
\label{fig:AVR_exponent}
\end{figure}

% If mass-dependent heating is strongly affecting our data, the exponent of the
% AVR should increase with decreasing effective temperature, \ie\ the heating
% rate should be greater for lower-mass stars.
% Figure \ref{fig:AVR_exponent} shows the AVR exponent does indeed increase with
% decreasing effective temperature, however, the AVR exponent calculated using
% more massive stars from the Geneva Copenhagen Survey (GCS)
% \citep{holmberg2009} is also shown on this plot as a dashed horizontal line.
% Most GCS stars are F and G type: 1-3 times more massive than the K dwarfs used
% in our study.
% If mass-dependent orbital heating is the main cause the rise in velocity
% dispersion with decreasing \teff\ seen in figure \ref{fig:age_cut}, then the
% AVR exponent of the more massive GCS stars should fall {\it below} the AVR
% exponent for the most massive stars in our sample.
% In other words, the dynamical heating rate should be lowest for the highest
% mass stars and highest for the lowest mass stars.
% % nordstrom2004, jorgensen2005,
% These AVR exponents were calculated with different populations of stars,
% subject to different selection effects, so directly comparing them comes with
% some risk.
% The most important difference is that the GCS AVR is calculated using \vz,
% however our AVRs are calculated using \vb\ since most stars in our sample do
% not have radial velocities.
% The median galactic latitude of these \kepler\ stars is 12.3\degrees, with
% latitudes ranging from 5.5\degrees\ to 21.4\degrees, so \vb\ is a relatively
% close approximation to \vz.
% Only 290 out of the 6820 stars in our sample, plotted in color in figure
% \ref{fig:age_cut}, have \gaia\ radial velocities, however we used the ones
% that do to compare the \vz\ AVR to the \vb\ AVR, across all temperatures
% between 3500 and 5000 K.
% For stars with gyrochronal ages between 0.5 and 4.5 Gyr, we measured a \vz\
% AVR exponent of 0.519 $\pm$ ... and a \vb\ AVR exponent of 0.514 $\pm$... .
% The similarity of these two AVR exponents suggests that directly comparing
% these two quantities {\it may} be a reasonable approach.
% For now, given the large difference between the (\vz) AVR exponent of the GCS
% sample and the (\vb) AVR exponent of the hottest stars in our sample, we
% assume that the increased velocity dispersion at cooler temperatures is mostly
% caused by incorrect age-grouping due to an incorrect period-color relation at
% old ages, and that any mass-dependent heating, while it may contribute at a
% low level to this result, is not the dominant driver.
% However, differentiating the effects of mass-dependent heating and the shape
% of the gyrochronology relations is certainly warranted in a follow-up study.
% }

The consistent velocity dispersion of young stars as a function of temperature
in figure \ref{fig:age_cut} shows that the Praesepe-calibrated gyrochronology
relation accurately predicts the relative ages of {\it young} field stars.
To our knowledge, no gyrochronology relation has never been demonstrated to
correctly predict ages (either relative or absolute) for such cool or such
young field stars.
This is not particularly remarkable however, since these young stars are a
similar age to the Praesepe cluster, which was used to calibrate the
period-color relation.

% Figure \ref{fig:age_comparison} shows the ages of star groups, predicted with
% the \citet{angus2019} gyrochronology relation, compared with the ages of star
% groups, predicted with the \citet{holmberg2009} AVR.
% \citet{holmberg2009} only provide an exponent (0.53), not an intercept, for the
% relation between logarithmic age (in Gyr) and logarithmic, \vz\ velocity
% dispersion (in \kms), so we fit the intercept to the mean \vb\ velocity
% dispersion across all temperatures, of our sample.
% \begin{figure}
%   \caption{
%       Ages predicted by the \citet{holmberg2009} \vz\ AVR, against ages
%     predicted by the \citet{angus2019} gyrochronology relation, for stars in
%     different temperature ranges.
% Points are colored by the effective temperature in the center of the \teff\
% bin.
% The dashed line shows the y=x relation.
% Predicted kinematic ages fan out over gyrochronal time, either because the AVR
%     is temperature-dependent or because the gyrochronology relations have an
%     age-dependent period-color relation.
% }
%   \centering
%     \includegraphics[width=1\textwidth]{age_comparison}
% \label{fig:age_comparison}
% \end{figure}
% Figure \ref{fig:age_comparison} shows a comparison between ages predicted by
% the \citet{holmberg2009} \vz\ AVR (using \vb\ as a proxy for \vz) and ages
% predicted by the \citet{angus2019} gyrochronology relation, for stars in
% different temperature ranges.
% The predicted kinematic ages fan out over gyrochronal time, either because the
% AVR is temperature-dependent or because the gyrochronology relations have an
% age-dependent period-color relation.
% Figure \ref{fig:age_comparion} also suggests that dynamical heating does not
% begin until after around ... Gyr:

% Alternatively, the ages of these young stars could be under-predicted by
% gyrochronology.
% These young stars have similar rotation periods to the Praesepe cluster, which
% was used to calibrate the gyrochronology relation used in this analysis.
% The ages of these young stars should, therefore, be the most accurate of the
% entire sample.
% However, they are tied to the age of the Praesepe cluster, whose age is not
% accurately known.
% \citet{angus2019} adopted an age for Praesepe of 650 Myrs, however the
% kinematic age-prediction for these stars is around 400 Myrs.
% Other studies have assigned a much older age of 800 Myrs to Praesepe
% \citep{brandt2015}.
% If the age assumed for Praesepe in the \citet{angus2019} gyrochronology
% calibration was inaccurate, the entire age {\it scale} would be incorrect.
% However, this is not what figure \ref{fig:age_comparion} shows -- it shows
% that the {\it relative} age of these youngest stars are under-predicted by
% kinematics or over-predicted by gyrochronology.

For now, we assume that the rise in velocity dispersions at cooler effective
temperatures is caused by an inaccurate period-color relation rather than
mass-dependent dynamical heating.
This assumption allows us to estimate what the shape of the period-color
relations {\it should} be.
Figure \ref{fig:age_cut} indicates that the period-color relation flattens
out, so we applied the same analysis to groups of stars with similar rotation
periods, equivalent to a completely flat period-color relation.
The top panel of figure \ref{fig:period_cut} shows the \mct\ sample with stars
in different period ranges plotted in different colors.
The bottom panel shows the velocity dispersion of each group as a function of
effective temperature.
\begin{figure}
  \caption{
This figure is similar to figure \ref{fig:age_cut}, with stars divided into
    period groups rather than age groups.
The velocity dispersion is more constant across effective temperatures for the
    most slowly rotating stars, compared to the stars selected with the
    \citet{angus2019} gyrochronology model, indicating that the gyrochronology
    models flatten out, and possibly even invert, at old ages.
}
  \centering
    \includegraphics[width=1\textwidth]{period_cut}
\label{fig:period_cut}
\end{figure}
Once again, the velocity dispersion increases with rotation period overall.
For the most rapidly rotating groups of stars, velocity dispersion decreases
with \teff\ as expected given the positively sloped period-color relation of
Praesepe and other young clusters: late K dwarfs rotate more slowly than early
K dwarfs of the same age.
Between rotation periods of 15 and 25 days, the temperature dependence of the
velocity dispersion starts to disappear, indicating that the period-color
relation becomes flat: late K dwarfs rotate at the same rate as early K dwarfs
of the same age.
At long rotation periods, the velocity dispersion still increases with
effective temperature, although the increase is much more modest compared to
the oldest stars in figure \ref{fig:age_cut}.
This suggests that rotation period may {\it decrease} with effective
temperature, \ie\ late K dwarfs may rotate {\it more} rapidly than early K
dwarfs.
This would be a paradigm shift for gyrochronology, stellar spin-down rate is
thought to be directly tied to magnetic field strength.
The deeper convection zones of later type stars generate stronger magnetic
fields which {\it should} lead to more efficient angular momentum loss.
Unfortunately, we still cannot rule out the possibility that mass-dependent
heating is acting on these stars which could be responsible for some, if not
all, of the increased velocity dispersion at cooler effective temperatures.
